---
title: "Banana,US"
author: "Team 1"
date: "03/12/2024"

execute:
  echo: false
  
format: 
  html:
    math: true
    toc: true  # Enable the table of contents
    toc-title: "Contents"  # Set the TOC title
    toc-depth: 2  # Limit TOC depth level
    code-fold: true  # Allow code blocks to be collapsible
    code-tools: false  # Show tools for code blocks (like copy button)
    code-copy: true  # Enable copy button for code blocks
    theme: Materia  # Use the "Lux" Bootswatch theme
    smooth-scroll: true  # Enable smooth scrolling for TOC links
    embed-resources: true  # Embed external resources directly
---

```{r}
#| echo: false
#| message: false
#| warning: false

# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))

# install.packages('ggplot2')
# install.packages('ggthemes')
# install.packages("stargazer")
# install.packages("wooldridge")
# install.packages("knitr")
# install.packages("modelsummary")
# install.packages("naniar")
# install.packages("kableExtra")
# install.packages("lmtest")
# install.packages("tsibble")
# install.packages("fable")
# install.packages("lubridate")
# install.packages("forecast")
# install.packages("urca")
# install.packages("broom")
# install.packages("plm")
# install.packages("fabletools")
# install.packages("lubridate")
# install.packages("plotly")
# install.packages("fpp3")



library(ggplot2)
library(ggthemes)
library(stargazer)
library(wooldridge)
library(knitr)
library(dplyr)
library(car)
library(modelsummary)
library(readxl)
library(naniar)
library(kableExtra)
library(lmtest)
library(tidyr)
library(tsibble)
library(lubridate)
library(forecast)
library(urca)
library(broom)
library(plm)
library(readr)
library(fable)
library(fabletools)
library(lubridate)
library(plotly)
library(gganimate)
library(fpp3)

invisible(Sys.setlocale("LC_TIME", "en_US.UTF-8")) # Set months in English without output
```

```{r}

banana_tsibble <- suppressMessages(
  read_csv("Cleaned_Commodity_Prices.csv", show_col_types = FALSE) |>
    select(Year, Month, `Banana, US`) |>
    mutate(Date = yearmonth(paste(Year, Month, sep = "-"))) |>
    select(Date, `Banana, US`) |>                             
    as_tsibble(index = Date)
)
```
```{r}

summary <- tibble(
  Metric = c("Start Year", "End Year", "Total Observations", "Min Price", "Max Price", "Mean Price", "Standard Deviation"),
  Value = c(
    min(year(banana_tsibble$Date)),
    max(year(banana_tsibble$Date)),
    nrow(banana_tsibble),
    round(min(banana_tsibble$`Banana, US`, na.rm = TRUE), 2),
    round(max(banana_tsibble$`Banana, US`, na.rm = TRUE), 2),
    round(mean(banana_tsibble$`Banana, US`, na.rm = TRUE), 2),
    round(sd(banana_tsibble$`Banana, US`, na.rm = TRUE), 2)
  )
)

knitr::kable(summary, col.names = c("Metric", "Value"), caption = "Banana Prices Dataset")
```

Data
```{r}

# Create an interactive line plot with plotly
p <- ggplot(banana_tsibble, aes(x = Date, y = `Banana, US`)) +
  geom_line(color = "blue") +
  ggtitle("Banana Prices Over Time (Levels)") +
  xlab("Date") +
  ylab("Price (US $/kg)") +
  theme_minimal()

# Convert ggplot to an interactive plotly object
ggplotly(p) %>%
  layout(title = "Banana Prices Over Time",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Price (US $/kg)"))
```
### Historical Events Affecting Banana Prices

- **1973–1979**: Oil crises led to increased transportation costs, causing price fluctuations.
- **Late 1990s**: Weather events (El Niño/La Niña) and changing trade policies caused instability.
- **2005–2015**: Steep price increase due to rising production costs, hurricanes, and higher global demand.
- **2022**: Sharp price drop driven by COVID-19 supply chain disruptions, inflation, and possible overproduction.

The banana market is influenced by weather conditions, geopolitical factors, pest outbreaks, and supply chain dynamics.
```{r}
# Create new columns for logs and first differences
banana_tsibble <- banana_tsibble %>%
  mutate(
    Log_Price = log(`Banana, US`),
    Diff_Price = difference(`Banana, US`)
  )

# Log-transformed prices
ggplot(banana_tsibble, aes(x = Date, y = Log_Price)) +
  geom_line(color = "darkgreen") +
  ggtitle("Log of Banana Prices Over Time") +
  xlab("Date") +
  ylab("Log Price (Log US $/kg)") +
  theme_minimal()

banana_tsibble_diff <- banana_tsibble %>% filter(!is.na(Diff_Price))

# Plot first differences
ggplot(banana_tsibble_diff, aes(x = Date, y = Diff_Price)) +
  geom_line(color = "red") +
  ggtitle("First Differences of Banana Prices Over Time") +
  xlab("Date") +
  ylab("Price Difference (US $/kg)") +
  theme_minimal()
```

Time Plot description

```{r}
banana_tsibble |>
  mutate(Month = month(Date, label = TRUE),   # Extract month as a factor
         Year = year(Date),                   # Extract year
         Decade = floor(Year / 10) * 10) |>   # Create a decade variable
  filter(Year %% 3 == 0) |>                   # Select only years divisible by 3
  ggplot(aes(x = Month, y = `Banana, US`, color = factor(Year), group = Year)) +
  geom_line() +
  labs(y = "$ (USD)",
       title = "Seasonal Plot: Banana Prices in US by Decade (Sampled Years)") +
  facet_wrap(~ Decade, scales = "fixed", ncol = 2) +  # Two columns for better readability
  theme_minimal() +
  theme(legend.position = "none",               # Remove the legend
        axis.text.x = element_text(angle = 45, hjust = 1))
```

Insights
```{r}
# Histogram and Kernel Density for Levels
ggplot(banana_tsibble, aes(x = `Banana, US`)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  geom_density(alpha = 0.4, fill = "blue") +
  ggtitle("Histogram and Kernel Density of Banana Prices (Levels)") +
  xlab("Price (US $/kg)") +
  ylab("Density") +
  theme_minimal()

# Histogram and Kernel Density for Logs
ggplot(banana_tsibble, aes(x = log(`Banana, US`))) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightgreen", color = "black") +
  geom_density(alpha = 0.4, fill = "green") +
  ggtitle("Histogram and Kernel Density of Log-Banana Prices") +
  xlab("Log Price (Log US $/kg)") +
  ylab("Density") +
  theme_minimal()
```
```{r}
# Create lagged values
banana_lagged <- banana_tsibble %>%
  mutate(Lag_1 = lag(`Banana, US`, 1)) %>% # Lag by 1 period
  filter(!is.na(Lag_1))                   # Remove NA values caused by lagging

# Scatter plot of levels vs lagged levels
ggplot(banana_lagged, aes(x = Lag_1, y = `Banana, US`)) +
  geom_point(color = "blue") +
  ggtitle("Scatter Plot: Banana Prices vs Lagged Banana Prices") +
  xlab("Lagged Price (US $/kg)") +
  ylab("Price (US $/kg)") +
  theme_minimal()
```
```{r}
# Calculate first differences of the time series
banana_diff <- banana_tsibble %>%
  mutate(Diff_Price = difference(`Banana, US`)) %>%
  filter(!is.na(Diff_Price))

ggAcf(banana_diff$Diff_Price) +
  ggtitle("Autocorrelation Function (ACF) of First Differences") +
  ylab("ACF") +
  xlab("Lag") +
  theme_minimal()
```


```{r}
banana_tsibble |>
  mutate(Month = month(Date, label = TRUE)) |>  # Extract Month as factor
  ggplot(aes(x = Month, y = `Banana, US`)) +
  geom_boxplot(fill = "lightblue", alpha = 0.6) +  # Boxplot
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "blue", linewidth = 1) +  # Mean line
  labs(y = "$/kg", 
       title = "Boxplot of Banana Prices by Month with Mean Line",
       x = "Month") +
  theme_minimal()
```

## Insights from the Boxplot:

1.  **Price Stability**:

    -   Most boxes have a similar size and position, indicating relatively stable banana prices throughout the year.
    -   There are no sharp changes in medians between months, meaning average banana prices remain at a similar level year-round.

2.  **Price Variability**:

    -   The whiskers are quite long, showing significant variability in prices across different years.
    -   The highest variability is observed in **January, June, and September**, as their whiskers are the longest, and there are several outliers.

3.  **Outliers**:

    -   Outliers appear in several months, especially in **January, June, and December**, which may indicate extreme banana prices in specific years.
    -   These outliers might be caused by economic or climatic events affecting banana supply during these periods.
```{r}
banana_tsibble %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  ggplot(aes(x = year(Date), y = `Banana, US`, color = Month)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linewidth = 1) +
  facet_wrap(~ Month, scales = "free_y") +
  labs(
    title = "Subseries Plot: Banana Prices by Month (Improved)",
    x = "Year",
    y = "Price (US $/kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
### Comments on Subseries Plot

- Banana prices show an **upward trend** across all months.
- **Higher variability** is observed in recent years.
- Prices exhibit a **consistent seasonal pattern** with peaks occurring around similar times each year.
