---
title: "Banana,US"
author: "Team 1"
date: "03/12/2024"

execute:
  echo: false
  
format: 
  html:
    math: true
    toc: true  # Enable the table of contents
    toc-title: "Contents"  # Set the TOC title
    toc-depth: 2  # Limit TOC depth level
    code-fold: true  # Allow code blocks to be collapsible
    code-tools: false  # Show tools for code blocks (like copy button)
    code-copy: true  # Enable copy button for code blocks
    theme: Materia  # Use the "Lux" Bootswatch theme
    smooth-scroll: true  # Enable smooth scrolling for TOC links
    embed-resources: true  # Embed external resources directly
---

```{r}
#| echo: false
#| message: false
#| warning: false

# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))

# install.packages('ggplot2')
# install.packages('ggthemes')
# install.packages("stargazer")
# install.packages("wooldridge")
# install.packages("knitr")
# install.packages("modelsummary")
# install.packages("naniar")
# install.packages("kableExtra")
# install.packages("lmtest")
# install.packages("tsibble")
# install.packages("fable")
# install.packages("lubridate")
# install.packages("forecast")
# install.packages("urca")
# install.packages("broom")
# install.packages("plm")
# install.packages("fabletools")
# install.packages("lubridate")
# install.packages("plotly")
# install.packages("fpp3")



library(ggplot2)
library(ggthemes)
library(stargazer)
library(wooldridge)
library(knitr)
library(dplyr)
library(car)
library(modelsummary)
library(readxl)
library(naniar)
library(kableExtra)
library(lmtest)
library(tidyr)
library(tsibble)
library(lubridate)
library(forecast)
library(urca)
library(broom)
library(plm)
library(readr)
library(fable)
library(fabletools)
library(lubridate)
library(plotly)
library(gganimate)
library(fpp3)

invisible(Sys.setlocale("LC_TIME", "en_US.UTF-8")) # Set months in English without output
```

```{r}

banana_tsibble <- suppressMessages(
  read_csv("Cleaned_Commodity_Prices.csv", show_col_types = FALSE) |>
    select(Year, Month, `Banana, US`) |>
    mutate(Date = yearmonth(paste(Year, Month, sep = "-"))) |>
    select(Date, `Banana, US`) |>                             
    as_tsibble(index = Date)
)
```
```{r}

summary <- tibble(
  Metric = c("Start Year", "End Year", "Total Observations", "Min Price", "Max Price", "Mean Price", "Standard Deviation"),
  Value = c(
    format(as.integer(min(year(banana_tsibble$Date))), nsmall = 0),
    format(as.integer(max(year(banana_tsibble$Date))), nsmall = 0),
    format(nrow(banana_tsibble), nsmall = 0),
    round(min(banana_tsibble$`Banana, US`, na.rm = TRUE), 2),
    round(max(banana_tsibble$`Banana, US`, na.rm = TRUE), 2),
    round(mean(banana_tsibble$`Banana, US`, na.rm = TRUE), 2),
    round(sd(banana_tsibble$`Banana, US`, na.rm = TRUE), 2)
  )
)

knitr::kable(summary, col.names = c("Metric", "Value"), caption = "Banana Prices Dataset")
```
> The World Bank compiled and published US banana price data in Commodity Markets Outlook. 
These prices are in US dollars per kilogram, represent the average market rates from global banana trade.
```{r}

# Create an interactive line plot with plotly
p <- ggplot(banana_tsibble, aes(x = Date, y = `Banana, US`)) +
  geom_line(color = "blue") +
  ggtitle("Banana Prices Over Time (Levels)") +
  xlab("Date") +
  ylab("Price (US $/kg)") +
  theme_minimal()

# Convert ggplot to an interactive plotly object
ggplotly(p) %>%
  layout(title = "Banana Prices Over Time",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Price (US $/kg)"))
```
### Historical Events Affecting Banana Prices

- **1973–1979**: Oil crises led to increased transportation costs, causing price fluctuations.
- **Late 1990s**: Weather events (El Niño/La Niña) and changing trade policies caused instability.
- **2005–2015**: Steep price increase due to rising production costs, hurricanes, and higher global demand.
- **2022**: Sharp price drop driven by COVID-19 supply chain disruptions, inflation, and possible overproduction.
- **2024**: Post COVID restore.

The banana market is influenced by weather conditions, geopolitical factors, pest outbreaks, and supply chain dynamics.
```{r}
# Create new columns for logs and first differences
banana_tsibble <- banana_tsibble %>%
  mutate(
    Log_Price = log(`Banana, US`),
    Diff_Price = difference(`Banana, US`)
  )

# Log-transformed prices
ggplot(banana_tsibble, aes(x = Date, y = Log_Price)) +
  geom_line(color = "darkgreen") +
  ggtitle("Log of Banana Prices Over Time") +
  xlab("Date") +
  ylab("Log Price (Log US $/kg)") +
  theme_minimal()

banana_tsibble_diff <- banana_tsibble %>% filter(!is.na(Diff_Price))

# Plot first differences
ggplot(banana_tsibble_diff, aes(x = Date, y = Diff_Price)) +
  geom_line(color = "red") +
  ggtitle("First Differences of Banana Prices Over Time") +
  xlab("Date") +
  ylab("Price Difference (US $/kg)") +
  theme_minimal()
```

> Log of Banana Prices Over Time Сhart reveals a long-term upward trend in banana prices, indicating consistent growth in proportional rather than nominal changes.<br>

>First Differences of Banana Prices Over Time reveals significant volatility in price changes over time, with notable spikes around the 2000s and 2020. The fluctuations become more pronounced as time progresses, suggesting an increase in the variability of price adjustments in the later years.

```{r}
banana_tsibble |>
  mutate(Month = month(Date, label = TRUE),   # Extract month as a factor
         Year = year(Date),                   # Extract year
         Decade = floor(Year / 10) * 10) |>   # Create a decade variable
  filter(Year %% 3 == 0) |>                   # Select only years divisible by 3
  ggplot(aes(x = Month, y = `Banana, US`, color = factor(Year), group = Year)) +
  geom_line() +
  labs(y = "$ (USD)",
       title = "Seasonal Plot: Banana Prices in US by Decade (Sampled Years)") +
  facet_wrap(~ Decade, scales = "fixed", ncol = 2) +  # Two columns for better readability
  theme_minimal() +
  theme(legend.position = "none",               # Remove the legend
        axis.text.x = element_text(angle = 45, hjust = 1))
```

> The seasonal plot shows that banana prices in the US exhibit minimal seasonal variation across decades, with a slight tendency for prices to peak around mid-year in some decades, such as the 2000s and 2020s. Overall, the trend reflects a relatively stable price pattern within each year, likely due to consistent supply and demand dynamics for bananas.

```{r}
# Histogram and Kernel Density for Levels
ggplot(banana_tsibble, aes(x = `Banana, US`)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "lightblue", color = "black") +
  geom_density(alpha = 0.4, fill = "blue") +
  ggtitle("Histogram and Kernel Density of Banana Prices (Levels)") +
  xlab("Price (US $/kg)") +
  ylab("Density") +
  theme_minimal()
```
> The histogram and kernel density plot reveal that banana prices in the US are positively skewed, with the majority of prices clustering below $0.5 per kilogram. However, there is a long tail extending toward higher price levels, indicating occasional periods of elevated prices that are less frequent.

```{r}
# Histogram and Kernel Density for Logs
ggplot(banana_tsibble, aes(x = log(`Banana, US`))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "lightgreen", color = "black") +
  geom_density(alpha = 0.4, fill = "green") +
  ggtitle("Histogram and Kernel Density of Log-Banana Prices") +
  xlab("Log Price (Log US $/kg)") +
  ylab("Density") +
  theme_minimal()
```
> The histogram and kernel density plot of the logarithm of banana prices exhibit a multimodal distribution, indicating distinct clusters of price levels over time. This transformation reduces skewness and highlights the presence of multiple regimes or patterns in banana pricing, potentially linked to market structure changes or external shocks.

```{r}
# Ensure banana_tsibble is correctly formatted as a tsibble
banana_tsibble <- banana_tsibble %>%
  as_tsibble(index = Date)

# Create lagged values
banana_lagged <- banana_tsibble %>%
  mutate(Lag_1 = dplyr::lag(`Banana, US`, 1)) %>%  # Use dplyr::lag for clarity
  filter(!is.na(Lag_1))                            # Remove NA values caused by lagging

# Scatter plot of levels vs lagged levels
ggplot(banana_lagged, aes(x = Lag_1, y = `Banana, US`)) +
  geom_point(color = "blue", alpha = 0.5) +        # Add transparency for better visualization
  ggtitle("Scatter Plot: Banana Prices vs Lagged Banana Prices") +
  xlab("Lagged Price (US $/kg)") +
  ylab("Price (US $/kg)") +
  theme_minimal()
```
> The scatter plot of banana prices versus lagged prices demonstrates a strong positive linear relationship, indicating high persistence in price levels over time. This suggests that banana prices exhibit significant autocorrelation, with current prices being highly dependent on their immediate past values. 

```{r}
# Calculate first differences of the time series
banana_diff <- banana_tsibble %>%
  mutate(Diff_Price = difference(`Banana, US`)) %>%
  filter(!is.na(Diff_Price))

ggAcf(banana_diff$Diff_Price) +
  ggtitle("Autocorrelation Function (ACF) of First Differences") +
  ylab("ACF") +
  xlab("Lag") +
  theme_minimal()
```
> The ACF plot of the first differences indicates that there is little significant autocorrelation at most lags, suggesting that the differencing has effectively removed the trend and made the series closer to stationarity. However, a few significant spikes at specific lags suggest some short-term dependencies or patterns still present in the data.

```{r}
banana_tsibble |>
  mutate(Month = month(Date, label = TRUE)) |>  # Extract Month as factor
  ggplot(aes(x = Month, y = `Banana, US`)) +
  geom_boxplot(fill = "lightblue", alpha = 0.6) +  # Boxplot
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "blue", linewidth = 1) +  # Mean line
  labs(y = "$/kg", 
       title = "Boxplot of Banana Prices by Month with Mean Line",
       x = "Month") +
  theme_minimal()
```
> The boxplot of banana prices by month shows relatively consistent median prices throughout the year, with minor seasonal variation as the mean line slightly decreases from February to September. The presence of outliers across months indicates occasional price volatility, but overall, monthly price distributions appear stable.
## Insights from the Boxplot:

1.  **Price Stability**:

    -   Most boxes have a similar size and position, indicating relatively stable banana prices throughout the year.
    -   There are no sharp changes in medians between months, meaning average banana prices remain at a similar level year-round.

2.  **Price Variability**:

    -   The whiskers are quite long, showing significant variability in prices across different years.
    -   The highest variability is observed in **January, June, and September**, as their whiskers are the longest, and there are several outliers.

3.  **Outliers**:

    -   Outliers appear in several months, especially in **January, June, and December**, which may indicate extreme banana prices in specific years.
    -   These outliers might be caused by economic or climatic events affecting banana supply during these periods.
```{r}
banana_tsibble %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  ggplot(aes(x = year(Date), y = `Banana, US`, color = Month)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linewidth = 1) +
  facet_wrap(~ Month, scales = "free_y") +
  labs(
    title = "Subseries Plot: Banana Prices by Month (Improved)",
    x = "Year",
    y = "Price (US $/kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
### Comments on Subseries Plot

- Banana prices show an **upward trend** across all months.
- **Higher variability** is observed in recent years.
- Prices exhibit a **consistent seasonal pattern** with peaks occurring around similar times each year.
